---
title: "Bulk RNA-seq Analysis — GSE168137 (5xFAD Alzheimer Mouse Model)"
subtitle: "Systematic phenotyping across cortex, hippocampus, and four timepoints"
author: "SLopezBegines"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-float: true
    code-fold: true
    code-tools: true
    embed-resources: true
    theme: cosmo
    fig-width: 10
    fig-height: 6
    fig-dpi: 150
execute:
  warning: false
  message: false
  cache: true
---

# Overview

This notebook analyses **GSE168137**, a comprehensive bulk RNA-seq dataset characterising
the **5xFAD mouse model** of Alzheimer's disease (Forner et al., *Scientific Data* 2021).

**Experimental design:**

| Factor    | Levels                              |
|-----------|-------------------------------------|
| Genotype  | 5xFAD (transgenic) vs BL6 (WT)     |
| Tissue    | Cortex, Hippocampus                 |
| Age       | 4, 8, 12, 18 months                 |
| Sex       | Female, Male                        |
| Samples   | 192 total                           |

The analysis follows five modular stages: (1) data import, (2) quality control,
(3) normalisation and filtering, (4) dimensionality reduction (PCA + UMAP), and
(5) differential expression with DESeq2 and edgeR. Results from both DE methods
are directly compared.

---

# 1. Setup

## 1.1 Parameters

Adjust these variables to subset the experiment before running the notebook.
Set `NULL` to keep all levels of that factor.

```{r params}
# ── Analysis parameters ──────────────────────────────────────────────────────
# Set to NULL to keep all; use character/integer vectors to subset

TISSUE_FILTER <- NULL           # "cortex" | "hippocampus" | NULL
AGE_FILTER    <- NULL               # e.g. c(4, 8) | NULL (all timepoints)
SEX_FILTER    <- NULL               # "Female" | "Male" | NULL

# DE contrast: always 5xFAD vs BL6
CONTRAST      <- c("genotype", "5xFAD", "BL6")

# DESeq2 design (covariates on top of genotype)
# If subsetting to a single tissue, drop tissue from the formula
DE_DESIGN     <- ~ sex + age_months + genotype

# LFC shrinkage method: "apeglm" (recommended) | "ashr"
SHRINK_TYPE   <- "apeglm"

# Output directory (relative to project root)
output_path   <- "output"
```

## 1.2 Packages and helper functions

```{r setup}
#| cache: false
source("../code/00_packages.R")
source("../code/01_global_variables.R")
source("../code/02_utils.R")
source("../code/03_import.R")
source("../code/04_qc.R")
source("../code/05_normalization.R")
source("../code/06_dimreduction.R")
source("../code/07_de_analysis.R")

# Ensure output directories exist
create_directories(output_path)

# Reset figure counter
image_number <- 1
```

---

# 2. Data Import

```{r import}
# Load raw counts and auto-parse sample metadata from column names
data_raw <- load_counts(
  count_file    = "../rawdata/GSE168137_countList.txt.gz",
  strip_version = TRUE
)

counts_raw <- data_raw$counts
meta_all   <- data_raw$metadata

# Preview metadata
knitr::kable(head(meta_all[, c("sample","genotype","tissue","age_months","sex")], 10),
             caption = "Sample metadata (first 10 rows)") |> kableExtra::kable_paper()
```

## Add gene names from Ensemble ID

```{r}
ensembl_ids <- rownames(counts_raw)

library(AnnotationDbi)
library(org.Mm.eg.db)

# Build Ensembl ID -> gene symbol mapping table
gene_map <- AnnotationDbi::select(
  org.Mm.eg.db,
  keys    = ensembl_ids,
  columns = c("ENSEMBL", "SYMBOL"),
  keytype = "ENSEMBL"
)

# Fall back to original Ensembl ID for any unmapped genes
gene_map$SYMBOL[is.na(gene_map$SYMBOL)] <- gene_map$ENSEMBL[is.na(gene_map$SYMBOL)]
# Apply mapping to counts matrix
rownames(counts_raw) <- gene_map$SYMBOL[match(rownames(counts_raw), gene_map$ENSEMBL)] 
```


```{r import-subset}
# Subset to selected tissue / age / sex
sub_data <- subset_experiment(
  counts   = counts_raw,
  metadata = meta_all,
  tissue   = TISSUE_FILTER,
  age      = AGE_FILTER,
  sex      = SEX_FILTER
)

counts_sub <- sub_data$counts
meta_sub   <- sub_data$metadata

cat(sprintf("\nAnalysis subset: %d genes x %d samples\n",
            nrow(counts_sub), ncol(counts_sub)))
cat(sprintf("Genotypes: %s\n",  paste(unique(meta_sub$genotype),   collapse=", ")))
cat(sprintf("Tissues:   %s\n",  paste(unique(meta_sub$tissue),     collapse=", ")))
cat(sprintf("Ages:      %s\n",  paste(sort(unique(meta_sub$age_months)), collapse=", ")))
cat(sprintf("Sexes:     %s\n",  paste(unique(meta_sub$sex),        collapse=", ")))
```

**Dataset:** The 5xFAD model carries five familial AD mutations and displays
progressive amyloid pathology. Wild-type C57BL/6J littermates serve as controls.
The count matrix contains raw read counts (integer, gene-level) aligned to
GRCm38 with Ensembl gene IDs (version suffix stripped).

---

# 3. Quality Control

## 3.1 Per-sample QC metrics

```{r qc-compute}
qc_df <- compute_qc_metrics(counts_sub, meta_sub)

# Print samples with low library size or low gene detection
qc_df |> filter(low_depth | low_coverage) |> dplyr::select(sample, low_depth, low_coverage) |> knitr::kable(
  caption = paste0("Samples flagged for low library size (< ",QC_MIN_LIBRARY_SIZE," reads) or low gene detection (< ", QC_MIN_GENES_DETECTED," genes).")) |> kableExtra::kable_paper()
# Save QC table
save_table(qc_df, "sample_QC_metrics", subdir = "QC")
```

```{r qc-library-size, fig.cap="Library sizes (total read counts) per sample. Samples below the dashed line (1 M reads) are flagged. Transparency indicates low-depth samples."}
p_lib <- plot_library_size(qc_df)
print(p_lib)
save_plot("library_size", p_lib, width = 14, height = 6, subdir = "QC")
```

```{r qc-genes, fig.cap="Number of genes detected per sample (count > 0), grouped by genotype × tissue and faceted by age."}
p_genes <- plot_genes_detected(qc_df)
print(p_genes)
save_plot("genes_detected", p_genes, width = 12, height = 5, subdir = "QC")
```

```{r qc-dist, fig.cap="Density of log10(library size) per group. Outlier samples appear as isolated peaks."}
p_dist <- plot_library_distribution(qc_df)
print(p_dist)
save_plot("library_distribution", p_dist, width = 12, height = 4, subdir = "QC")
```

## 3.2 QC summary

```{r qc-summary}
flagged <- qc_summary(qc_df)
```

```{r qc-table}
#| echo: false
knitr::kable(
  qc_df[, c("sample","genotype","tissue","age_months","sex",
             "library_size","genes_detected","low_depth","low_coverage")],
  caption = "QC metrics per sample"
) |> kableExtra::kable_paper()
```

## 3.3 Sample correlation

Inter-sample Pearson correlation on the top 1,000 most variable genes (log2 CPM).
Samples should cluster by tissue and then genotype.

```{r qc-correlation, fig.cap="Pearson correlation heatmap. Ideal bulk RNA-seq samples cluster by experimental group.", fig.width=10, fig.height=9}
# Using log2(CPM+1) for the correlation — no normalisation model required here
cpm_sub <- counts_to_cpm(counts_sub)

p_cor <- plot_sample_correlation(cpm_sub, meta_sub, top_n = 1000, log_transform = TRUE)

# Save as PDF (pheatmap returns a recordedplot)
pdf(file.path(output_path, "figures", "QC",
              sprintf("%03d_sample_correlation.pdf", image_number)),
    width = 10, height = 9)
print(p_cor)
dev.off()
image_number <- image_number + 1
```

---

# 4. Normalisation and Filtering

## 4.1 Gene-level pre-filtering

Remove genes that do not reach a minimum CPM in at least the number of samples
in the smallest group. This step reduces multiple-testing burden without
discarding biologically informative genes.

```{r filter}
counts_filt <- filter_low_counts(
  counts_sub,
  min_cpm     = MIN_COUNT_CPM,
  min_samples = MIN_SAMPLES_EXPRESSED,
  min_total   = MIN_TOTAL_COUNTS
)
```

## 4.2 Build DESeqDataSet and estimate size factors

```{r dds-build}
dds <- build_dds(counts_filt, meta_sub, design = DE_DESIGN)

# Estimate size factors (needed for VST and normalization plots)
dds <- DESeq2::estimateSizeFactors(dds)
```

```{r size-factors, fig.cap="DESeq2 size factors per sample. Values close to 1 indicate similar library composition. Large deviations suggest outlier samples or batch effects."}
p_sf <- plot_size_factors(dds, meta_sub)
print(p_sf)
save_plot("size_factors", p_sf, width = 14, height = 5, subdir = "normalization")
```

## 4.3 Variance-stabilising transformation (VST)

VST is applied for visualisation (PCA, UMAP, correlation heatmaps). Raw counts
are used for the actual DE tests.

```{r vst}
vst_obj <- normalize_vst(dds, blind = TRUE)

# Save checkpoint
save_checkpoint(vst_obj, "vst")
```

```{r mean-variance, fig.cap="Mean-SD plot after VST normalisation. Successful VST produces a flat lowess line, indicating variance is no longer correlated with mean expression."}
vsn::meanSdPlot(SummarizedExperiment::assay(vst_obj))
```

---

# 5. Dimensionality Reduction

## 5.1 PCA

PCA is computed on the `r N_TOP_GENES_PCA` most variable genes after VST normalisation.

```{r pca-run}
pca_res <- run_pca(vst_obj, n_top = N_TOP_GENES_PCA, n_pcs = N_PCS)
```

```{r scree, fig.cap="Scree plot: proportion of variance captured by each PC. The elbow indicates where signal transitions to noise."}
p_scree <- plot_scree(pca_res$var_explained, n_show = 20)
print(p_scree)
save_plot("PCA_scree", p_scree, width = 8, height = 5, subdir = "PCA")
```

```{r pca-grid, fig.cap="PCA coloured by experimental covariates. Separation along PC1/PC2 identifies the dominant sources of variation.", fig.width=12, fig.height=8}
p_pca_grid <- plot_embedding_grid(
  embed_df      = pca_res$pca_df,
  x_col         = "PC1", y_col = "PC2",
  var_explained = pca_res$var_explained
)
print(p_pca_grid)
save_plot("PCA_grid_covariates", p_pca_grid, width = 12, height = 8, subdir = "PCA")
```

```{r pca-genotype, fig.cap="PC1 vs PC2 coloured by genotype. Separation between 5xFAD and BL6 confirms the transcriptional effect of the transgene."}
p_pca_geno <- plot_pca(pca_res$pca_df, pca_res$var_explained,
                        colour_by = "genotype", shape_by = "sex",
                        palette = GENOTYPE_COLORS)
print(p_pca_geno)
save_plot("PCA_genotype", p_pca_geno, width = 7, height = 6, subdir = "PCA")
```

```{r pca-age, fig.cap="PC1 vs PC2 coloured by age. Age-related progression is visible if samples separate along a temporal axis."}
p_pca_age <- plot_pca(pca_res$pca_df, pca_res$var_explained,
                       colour_by = "age_months", shape_by = "genotype",
                       palette = AGE_COLORS)
print(p_pca_age)
save_plot("PCA_age", p_pca_age, width = 7, height = 6, subdir = "PCA")
```

## 5.2 UMAP

UMAP further separates the sample structure using the top `r N_PCS_UMAP` PCs.
A fixed seed ensures reproducibility.

```{r umap-run}
umap_df <- run_umap(pca_res, n_pcs = N_PCS_UMAP)
```

```{r umap-grid, fig.cap="UMAP coloured by experimental covariates.", fig.width=12, fig.height=8}
p_umap_grid <- plot_embedding_grid(umap_df, x_col = "UMAP1", y_col = "UMAP2")
print(p_umap_grid)
save_plot("UMAP_grid_covariates", p_umap_grid, width = 12, height = 8, subdir = "UMAP")
```

```{r umap-genotype, fig.cap="UMAP coloured by genotype."}
p_umap_geno <- plot_umap(umap_df, colour_by = "genotype", shape_by = "tissue",
                          palette = GENOTYPE_COLORS)
print(p_umap_geno)
save_plot("UMAP_genotype", p_umap_geno, width = 7, height = 6, subdir = "UMAP")
```

```{r umap-age, fig.cap="UMAP coloured by age (months). A trajectory from left to right reflects progressive disease."}
p_umap_age <- plot_umap(umap_df, colour_by = "age_months", shape_by = "genotype",
                         palette = AGE_COLORS)
print(p_umap_age)
save_plot("UMAP_age", p_umap_age, width = 7, height = 6, subdir = "UMAP")
```

---

# 6. Differential Expression — DESeq2

## 6.1 Model and contrast

We compare **5xFAD vs BL6** (reference), adjusting for sex and age as covariates.
A gene is called significant at padj < `r P_VAL_THRESH` and |log2FC| ≥ `r FC_THRESH`.

```{r deseq2-run}
#| cache: true
if (check_checkpoint("deseq2_res")) {
  deseq2_res <- load_checkpoint("deseq2_res")
} else {
  # Rebuild dds with design (not blind) for DE
  dds_de <- build_dds(counts_filt, meta_sub, design = DE_DESIGN)
  deseq2_res <- run_deseq2(dds_de, contrast = CONTRAST, shrink_type = SHRINK_TYPE)
  save_checkpoint(deseq2_res, "deseq2_res")
}

# Save full results table
save_table(deseq2_res, "DESeq2_5xFAD_vs_BL6", subdir = "DE")
```

## 6.2 Results summary

```{r deseq2-summary}
n_up   <- sum(deseq2_res$direction == "UP",   na.rm = TRUE)
n_down <- sum(deseq2_res$direction == "DOWN",  na.rm = TRUE)
cat(sprintf("DESeq2 significant DE genes:\n  UP:   %d\n  DOWN: %d\n  Total: %d\n",
            n_up, n_down, n_up + n_down))

knitr::kable(
  head(deseq2_res[deseq2_res$significant, c("gene","baseMean","log2FoldChange","lfcSE","padj","direction")], 20),
  caption = "Top 20 DESeq2 significant genes (ranked by padj)"
) |> kableExtra::kable_paper()
```

## 6.3 Volcano plot

```{r deseq2-volcano, fig.cap="DESeq2 volcano plot. Red: upregulated in 5xFAD; blue: downregulated. Dashed lines mark padj=0.05 and |log2FC|=1."}
p_volc_d2 <- plot_volcano(deseq2_res,
                            title   = "5xFAD vs BL6 — DESeq2",
                            lfc_col = "log2FoldChange", pval_col = "padj")
print(p_volc_d2)
save_plot("DESeq2_volcano", p_volc_d2, width = 8, height = 7, subdir = "DE/DESeq2")
```

## 6.4 MA plot

```{r deseq2-ma, fig.cap="DESeq2 MA plot. Genes with high average expression generally have well-estimated fold-changes. The apeglm shrinkage pulls extreme LFC estimates towards zero for lowly expressed genes."}
p_ma_d2 <- plot_ma(deseq2_res, title = "5xFAD vs BL6 — DESeq2 (apeglm shrinkage)",
                    lfc_col = "log2FoldChange", mean_col = "baseMean")
print(p_ma_d2)
save_plot("DESeq2_MA", p_ma_d2, width = 8, height = 6, subdir = "DE/DESeq2")
```

## 6.5 Heatmap of top DE genes

```{r deseq2-heatmap, fig.cap="Heatmap of top 50 DESeq2 DE genes (z-scored VST expression). Columns annotated by genotype, tissue, age, and sex.", fig.height=10}
# Rebuild VST with blind=FALSE for DE-focused heatmap
vst_de <- normalize_vst(build_dds(counts_filt, meta_sub, design = DE_DESIGN),
                         blind = FALSE)

p_hm_d2 <- plot_de_heatmap(deseq2_res, vst_de, meta_sub,
                             n_top = 50, title = "Top 50 DE genes — DESeq2")
save_plot("DESeq2_heatmap_top50", p_hm_d2, width = 12, height = 10, subdir = "DE/DESeq2")
```

---

# 7. Differential Expression — edgeR

## 7.1 Model and contrast

edgeR uses TMM normalisation and a quasi-likelihood F-test for the same
5xFAD vs BL6 contrast. We pass the same design formula to ensure comparability.

```{r edger-run}
#| cache: true
if (check_checkpoint("edger_res")) {
  edger_res <- load_checkpoint("edger_res")
} else {
  edger_res <- run_edger(
    counts         = counts_filt,
    metadata       = meta_sub,
    design_formula = DE_DESIGN,
    contrast_col   = "genotype",
    numerator      = "5xFAD",
    denominator    = "BL6"
  )
  save_checkpoint(edger_res, "edger_res")
}

save_table(edger_res, "edgeR_5xFAD_vs_BL6", subdir = "DE")
```

## 7.2 Results summary

```{r edger-summary}
n_up_er   <- sum(edger_res$direction == "UP",   na.rm = TRUE)
n_down_er <- sum(edger_res$direction == "DOWN",  na.rm = TRUE)
cat(sprintf("edgeR significant DE genes:\n  UP:   %d\n  DOWN: %d\n  Total: %d\n",
            n_up_er, n_down_er, n_up_er + n_down_er))

knitr::kable(
  head(edger_res[edger_res$significant, c("gene","logFC","logCPM","F","PValue","FDR","direction")], 20),
  caption = "Top 20 edgeR significant genes (ranked by PValue)"
)
```

## 7.3 Volcano plot

```{r edger-volcano, fig.cap="edgeR volcano plot. Same thresholds as DESeq2 for direct visual comparison."}
p_volc_er <- plot_volcano(edger_res,
                           title   = "5xFAD vs BL6 — edgeR",
                           lfc_col = "logFC", pval_col = "FDR")
print(p_volc_er)
save_plot("edgeR_volcano", p_volc_er, width = 8, height = 7, subdir = "DE/edgeR")
```

## 7.4 MA plot

```{r edger-ma, fig.cap="edgeR MA plot (logFC vs average logCPM across all samples)."}
p_ma_er <- plot_ma(edger_res, title = "5xFAD vs BL6 — edgeR",
                    lfc_col = "logFC", mean_col = "logCPM")
print(p_ma_er)
save_plot("edgeR_MA", p_ma_er, width = 8, height = 6, subdir = "DE/edgeR")
```

## 7.5 Heatmap of top DE genes

```{r edger-heatmap, fig.cap="Top 50 edgeR DE genes (z-scored VST expression).", fig.height=10}
p_hm_er <- plot_de_heatmap(edger_res, vst_de, meta_sub,
                             n_top = 50, title = "Top 50 DE genes — edgeR")
save_plot("edgeR_heatmap_top50", p_hm_er, width = 12, height = 10, subdir = "DE/edgeR")
```

---

# 8. DESeq2 vs edgeR Comparison

A key quality check: results from DESeq2 and edgeR should be highly concordant
for a well-powered dataset. Discordances often reflect differences in how each
method handles dispersion estimation at extreme count values.

## 8.1 Gene overlap

```{r compare-venn}
merged_de <- merge_de_results(deseq2_res, edger_res)
overlap    <- venn_de_overlap(merged_de)

save_table(merged_de, "DESeq2_edgeR_merged", subdir = "DE")
```

## 8.2 LFC correlation

```{r compare-lfc, fig.cap="Scatter plot of log2FC estimates from DESeq2 (x) vs edgeR (y). The diagonal dashed line represents perfect agreement. Pearson r is shown."}
p_lfc_comp <- plot_lfc_comparison(merged_de)
print(p_lfc_comp)
save_plot("comparison_LFC", p_lfc_comp, width = 7, height = 6, subdir = "DE/comparison")
```

## 8.3 Adjusted p-value ranks

```{r compare-pval, fig.cap="Rank-rank comparison of adjusted p-values. DESeq2 padj (x) vs edgeR FDR (y). Points near the diagonal indicate consistent gene-level rankings."}
p_pval_comp <- plot_pval_comparison(merged_de)
print(p_pval_comp)
save_plot("comparison_pval_ranks", p_pval_comp, width = 7, height = 6,
          subdir = "DE/comparison")
```

## 8.4 Side-by-side volcano plots

```{r compare-volcanos, fig.cap="Volcano plots from both methods side-by-side for direct comparison.", fig.width=14, fig.height=6}
(p_volc_d2 + ggtitle("DESeq2")) | (p_volc_er + ggtitle("edgeR"))
```

## 8.5 Concordant gene list

```{r concordant-genes}
concordant_genes <- merged_de$gene[
  !is.na(merged_de$agreement) & merged_de$agreement == "Concordant"]

cat(sprintf("Concordant significant genes: %d\n", length(concordant_genes)))

concordant_df <- merged_de[merged_de$agreement == "Concordant",
  c("gene", "log2FoldChange_deseq2", "padj_deseq2",
    "logFC_edger", "FDR_edger", "direction_deseq2")]
concordant_df <- concordant_df[order(concordant_df$padj_deseq2), ]

knitr::kable(
  head(concordant_df, 30),
  caption = "Top 30 concordant DE genes (ranked by DESeq2 padj)"
) |> kableExtra::kable_paper()

save_table(concordant_df, "concordant_DE_genes", subdir = "DE")
```

---

# 9. Session Information

```{r session-info}
#| cache: false
sessionInfo()
```
